---
title: "Analysis_script"
editor: visual
date: today
toc: false
abstract-title: "Research question:"
abstract: "Does ICS-X lead to a greater improvement of forced expiratory volume (FEV1) compared to a standard inhaler in patients with milkd and persistent asthma?"
format:
  html:
    
    embed-resources: true 
    number-sections: true 
    code-fold: show 
editor_options:
  chunk_output_type: console
execute:
  
  warning: false 
  message: false
---

## Introduction

A randomized controlled trial evaluates a new inhaled corticosteroid (ICS-X) compared to a standard inhaler in patients with mild persistent asthma.

Patients are randomized to receive either ICS-X or standard therapy for 6 weeks. Lung function is assessed at baseline and at follow-up using FEV₁ (forced expiratory volume in 1 second).

***The primary endpoint is improvement in FEV₁.***

### Variables

-   *Outcome:*

    fev1_improvement: change in FEV₁ (liters) from baseline to 6 weeks

-   *Continuous predictors:*

    -   age: patient age (years)
    -   baseline_fev1: baseline FEV₁ (liters)

-   Categorical predictors:

    -   treatment_group: randomized treatment ("standard_inhaler", "ICSX")

    -   smoking_status: smoking status ("never", "former", "current")

### Hypotheses:

-   ICS-X leads to greater improvement in FEV₁
-   Higher baseline FEV₁ is associated with less room for improvement
-   Smokers exhibit a reduced treatment response
-   A negative Age effect on improvement

## Data Import and Screening

```{r}
#| label: Importing-data
#| include: false
#| echo: false

rm(list = ls())

# Import necessary libraries
  library(performance)
  library(pastecs)
  library(car)
  library(ggplot2)
  library(ggpubr)
  library(ggeffects)
  library(GGally)
  library(tidyverse)
  library(glue)

# load data
dd <- read.csv("data/processed/lung-function.csv", stringsAsFactors = T)

# Screen data
str(dd)
summary(dd)

# Check for NA's
sum(is.na(dd)) # No NA

# Check balanced RCT design
table(dd$treatment_group)
```

Data was successfully imported and screened. No inconsistencies, further information in the data exploration sections.

## Numerical Exploration

```{r}
#| label: tbl-Numerical
#| tbl-cap: "Table showing various descriptive statistics of the Asthma RCT superiority trial"
#| echo: false
#| message: false
#| warning: false
#| paged-print: false

corr_matrix <- round(cor(sapply(dd, as.numeric)), 2)

table <- round(stat.desc(dd[,1:4], norm = T), 2)
table_final <- knitr::kable(table, align = "c", col.names = c("Patient ID",
                                                "FEV1 Improvment from Baseline",
                                                "Age of Patient",
                                                "Baseine FEV1"))

table_final
```

After importing and screening the data, it seems that everything is in order. The @tbl-Numerical shows the Patient ID, FEV1 improvement score, Age of the Patient and the Baseline FEV1. The last two predictors, mainly the treatment group and smoking status were not included, as their descriptive statistics are not interpret-able due to them being categorical variables. Additionally, in the numerical exploration a correlation matrix was done to explore possible pre-preemptive correlations as well as the possibility of co-linearity, a violation of our later model choice. The correlation matrix showed a weak to moderate negative correlation between the treatment group (ICSX or standard inhaler) with a value of `r corr_matrix[5,2]`\`. In other words, the ICSX treatment in this preemptive exploration accounts for a `r abs(corr_matrix[5,2])` higher FEV1 after intervention then the standard inhaler, however this is very early on in the process, meaning this value cannot be interpreted, instead gave us a good indication of a possible effect direction. Addtionally, it was intresting that the FEV1 improvement also heavily depended on the FEV1 baseline showing a moderate negative correlation of `r corr_matrix[4,2]`\`. The final variable that contained intresting information was the smoking status, showing a weak to moderate negative correlation of `r corr_matrix[6, 2]`\`with FEV1 improvement. This will need to be further explored graphically.

## Graphical Exploration

```{r}
#| label: fig-Graphical-Exploration
#| fig-cap: "Graph helping to visualize the research question"
#| echo: false
#| include: true


p1 <- ggplot() + 
  geom_point(data = dd, 
             aes(x = baseline_fev1, y = fev1_improvement, col = treatment_group)) + 
  geom_smooth(data = dd, 
             aes(x = baseline_fev1, y = fev1_improvement, col = treatment_group),
             method = "lm") + 
  theme_classic() + 
  xlab("Baseline FEV1 value") + 
  ylab("FEV1 Improvment from Baseline")

p2 <- ggplot() + 
  geom_point(data = dd, 
             aes(x = smoking_status, y = fev1_improvement, col = treatment_group)) + 
  geom_boxplot(data = dd, 
             aes(x = smoking_status, y = fev1_improvement, col = treatment_group)) + 
  theme_classic() + 
  xlab("Smoking Status") + 
  ylab("FEV1 Improvment from Baseline")

ggarrange(p1, p2, common.legend = T)
```

As shown in both the left and right plot in the @fig-Graphical-Exploration, there is a negative relationship for both treatment groups (standard treatment and ICSX) for increasing baseline FEV1 value. Although counterintuitive, the plot on the right showing smoking status vs FEV1 improvement indicates a positive relationship between increasing smoking and FEV1 improvement, nonetheless the confidence is low. Additionally, this makes me consider a interaction based model as this is biologically not as clear to me.

## Model Fitting and Selection

Since the outcome is continous, we will fit a linear model. Additionally, the assumption of normality is fullfilled as shown by the shapiro wilk test returning a p-value of `r shapiro.test(dd$fev1_improvement)[2]`\`. However, the full assumptions will be checked later.

Variables being considered :

-   **Outcome**: FEV1 improvment

-   **Predictors**: treatment, FEV1 baseline, smoking status

These variables are considered due to apriori understanding of respiratory function, as well as a initial graphical and numerical exploration indicating these variables contain a large association with the outcome.

```{r}
#| label: Fitting-models
#| echo: true
#| output: false


model.null <- lm(data = dd, fev1_improvement ~ 1)
model.1 <- lm(data = dd, fev1_improvement ~ treatment_group)
model.2 <- lm(data = dd, fev1_improvement ~ treatment_group + baseline_fev1)
model.3 <- lm(data = dd, fev1_improvement ~ treatment_group * baseline_fev1)
model.4 <- lm(data = dd, fev1_improvement ~ treatment_group + baseline_fev1 + smoking_status)
model.5 <- lm(data = dd, fev1_improvement ~ treatment_group + baseline_fev1 * smoking_status)
model.6 <- lm(data = dd, fev1_improvement ~ treatment_group * baseline_fev1 * smoking_status)

compare_models <- anova(model.null, model.1, model.2, model.3, model.4, model.5, model.6)

AIC_compare <- compare_performance(model.null, model.1, model.2, model.3, model.4, model.5, model.6)
```

I will go with model.4, which contains additivly the predictor variables `r colnames(dd)[6]`\`, `r colnames(dd)[4]`\` and of course our `r colnames(dd)[5]`\`. I chose this model both because it is biologically plausible, but also because it preforms better then the other models fitted. Since all models fitted were nested, a anova test under the F distribution allows us to compare which additionally parameter was worth it. In our case model 4 (chosen model), which contained all three predictors additively explained a total of `r compare_models[1, "RSS"] - compare_models[5, "RSS"]`\` out of the original unexplained `r compare_models[1, "RSS"]`\`. This was statistically significant with a F statistic of `r compare_models[5, "F"]`\`which corresponds to a p-value of \< `r format.pval(compare_models[5, "Pr(>F)"])`\`.

## Model Validation

```{r}
#| label: fig-Model-validation
#| echo: true
#| fig-cap: Check model function providing 5 graphs summarizing the model assumptions and whether or not they are fullfilled
#| fig-height: 16
#| fig-width: 10
#| output: true

model <- model.4

check_model(model)
```

As one can see the data is simulated as all the assumptions, show in @fig-Model-validation are not only fulfilled but awkwardly so. Nonetheless, there is no co linearity, there is homogeneous variance in the residuals, no outliers or high leverage points, the normality assumption is fulfilled and our models predicted data seems to match our real data quite well. No red flags, thus it is possible to move on to the summary and prediction plots.

## Model Summary

```{r}
#| label: fig-Prediction-Plot
#| fig-cap: "Figure showing the respective effect sizes for treatment controlling for smoking and FEV1 baseline"
#| echo: false
#| output: true

s <- seq(min(dd$baseline_fev1), max(dd$baseline_fev1), length.out = 100)

pred_data <- predict_response(model,
                              c("baseline_fev1 [s]",
                                "smoking_status",
                                "treatment_group"))

prediction_plot <- ggplot(dd) + 
  geom_point(data = dd, aes(x = baseline_fev1, y = fev1_improvement, col = treatment_group),
             position = position_jitter(width = .4)) + 
  facet_wrap(~smoking_status) + 
  geom_line(data = pred_data, 
            aes(x = x, y = predicted, col = facet)) + 
  geom_ribbon(data = pred_data, 
              aes(x= x, y = predicted, fill = facet, ymin = conf.low, ymax = conf.high),
              alpha = .2) +
  facet_wrap(~group) + 
  theme_bw() +
  xlab("Baseline FEV1 value for three levels of Smokers") + 
  ylab("FEV1 improvment from baseline after intervention") + 
  scale_color_discrete(
    name = "Treatment Group",
    labels = c("ICS-X", "Standard Inhaler")
  ) +
  scale_fill_discrete(
    name = "Treatment Group",
    labels = c("ICS-X", "Standard Inhaler")
  ) + 
  theme(legend.position = "bottom")

print(prediction_plot)

```

```{r}
#| label: tbl-summary
#| tbl-cap: "Table showing the model estimates, standard errors, statistical estimates and likely hood of them happening (p-value)"
#| echo: false
#| output: true


model_output <- summary(model)$coeff

knitr::kable(round(model_output, 3), align = "c")
```

Although hard to see and might have been a little over-the-top trying to include all predictors, @fig-Prediction-Plot shows the predicted effect of smoking status, treatment group and Baseline FEV1 value on the FEV1 improvement after intervention. I tried to change the row.names of the summary table shown in @tbl-summary however i kept getting errors, nonetheless... The intercept which has an estimate of `r model_output[1,1]`\`has no biological meaning, since this is the mean FEV1 improvement score for ICSX treatment for a current smoker, however with a baseline FEV1 of 0. Thus, the model mean FEV1 improvement of a curent smoker and ICSX treatment for a mean baseline FEV1 value is `r model_output[1,1] + mean(dd$baseline_fev1)`\`. As the person reduces their smoking from current to former and then never, we see a reduction in the FEV1 improvement score by `r model_output[4,1]`\` for former smokers and `r model_output[5,1]`\`if the patient never smoked. This is suprising to me and counterintuitive, however who knows, biology! In addition, to hammer the point home these values are statistically significant with a p-value of \> `r model_output[4,4]`\` and \>`r model_output[5,4]`\`respectively.

The interesting science however is the effect of the treatment group, this showed a negative effect, providing evidence for the superiority of the ICSX treatment compared to the standard treatment with an effect size of `r model_output[1,2]`\` with the confidence interval ranging from `r confint(model)[2]`\`.

I could keep writing but my hands hurt and i hope i showed that i understand the model output.

```{r}
#| label: version used
#| echo: false


version<- sessionInfo()
```
