---
title: "Analysis_script"
editor: visual
date: today
toc: false
abstract-title: "Research question:"
abstract: "Does ICS-X lead to a greater improvement of forced expiratory volume (FEV1) compared to a standard inhaler in patients with milkd and persistent asthma?"
format:
  html:
    
    embed-resources: true 
    number-sections: true 
    code-fold: show 
editor_options:
  chunk_output_type: console
execute:
  
  warning: false 
  message: false
---

## Introduction

A randomized controlled trial evaluates a new inhaled corticosteroid (ICS-X) compared to a standard inhaler in patients with mild persistent asthma.

Patients are randomized to receive either ICS-X or standard therapy for 6 weeks. Lung function is assessed at baseline and at follow-up using FEV₁ (forced expiratory volume in 1 second).

***The primary endpoint is improvement in FEV₁.***

### Variables

-   *Outcome:*

    fev1_improvement: change in FEV₁ (liters) from baseline to 6 weeks

-   *Continuous predictors:*

    -   age: patient age (years)
    -   baseline_fev1: baseline FEV₁ (liters)

-   Categorical predictors:

    -   treatment_group: randomized treatment ("standard_inhaler", "ICSX")

    -   smoking_status: smoking status ("never", "former", "current")

### Hypotheses:

-   ICS-X leads to greater improvement in FEV₁
-   Higher baseline FEV₁ is associated with less room for improvement
-   Smokers exhibit a reduced treatment response
-   A negative Age effect on improvement

## Data Import and Screening

```{r}
#| label: Importing-data
#| include: false
#| echo: false

rm(list = ls())

# Import necessary libraries
  library(performance)
  library(pastecs)
  library(car)
  library(ggplot2)
  library(ggpubr)
  library(ggeffects)
  library(GGally)
  library(tidyverse)
  library(glue)

# load data
dd <- read.csv("data/processed/lung-function.csv", stringsAsFactors = T)

# Screen data
str(dd)
summary(dd)

# Check for NA's
sum(is.na(dd)) # No NA

# Check balanced RCT design
table(dd$treatment_group)
```

Data was successfully imported and screened. No inconsistencies, further information in the data exploration sections.

## Numerical Exploration

```{r}
#| label: tbl-Numerical
#| tbl-cap: "This is a table"
#| echo: false
#| message: false
#| warning: false
#| paged-print: false

corr_matrix <- round(cor(sapply(dd, as.numeric)), 2)

table <- round(stat.desc(dd[,1:4], norm = T), 2)
table_final <- knitr::kable(table, align = "c", col.names = c("Patient ID",
                                                "FEV1 Improvment from Baseline",
                                                "Age of Patient",
                                                "Baseine FEV1"))

table_final
```

After importing and screening the data, it seems that everything is in order. The @tbl-Numerical shows the Patient ID, FEV1 improvement score, Age of the Patient and the Baseline FEV1. The last two predictors, mainly the treatment group and smoking status were not included, as their descriptive statistics are not interpret-able due to them being categorical variables. Additionally, in the numerical exploration a correlation matrix was done to explore possible pre-preemptive correlations as well as the possibility of co-linearity, a violation of our later model choice. The correlation matrix showed a weak to moderate negative correlation between the treatment group (ICSX or standard inhaler) with a value of `r corr_matrix[5,2]`\`. In other words, the ICSX treatment in this preemptive exploration accounts for a 0.2 higher FEV1 then the standard inhaler, however this is very early on in the process, meaning this value cannot be interpreted, instead gave us a good indication of a possible effect direction. Addtionally, it was intresting that the FEV1 improvement also heavily depended on the FEV1 baseline showing a moderate negative correlation of `r corr_matrix[4,2]`\`. The final variable that contained intresting information was the smoking status, showing a weak to moderate negative correlation of `r corr_matrix[6, 2]`\`with FEV1 improvement. This will need to be further explored graphically.\
